---
title: "ProgramEvalPS1"
author: "Jessica Prus and Arianna Wooten"
date: "2026-04-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stargazer)  
library(broom)      
library(kableExtra) 
library(tinytex)
```

```{r}
## load data
df_data <- read_csv("C:/Users/blueb/Downloads/ps1_data_spring26.csv", show_col_types = FALSE)
```

```{r}
glimpse(df_data)
summary(df_data)
```


1.


2. Measuring household specific treatment effects is impossible because of 
selection bias. The households who receive and don't receive the microcredit 
might be systematically different from each other, would would in turn affect 
the magnitude of the results which may not represent the true impact. 

3.


4. In this context, the ATT would be the average outcome from receiving the 
microcredit on households who did receive the microcredit subtracted by the 
average outcome from not receiving the microcredit on the households who did 
receive the microcredit. The FARMS data will not allow us to estimate the ATT, 
because we are unable to observe the outcomes of households who received the 
microcredit had they not received the microcredit. 


5.



6.


7.


8.


9.


10.



