---
title: "Problem Set 2"
author: "Jessica Prus and Arianna Wooten"
date: "2026-04-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(tidyverse)
library(stargazer)  
library(broom)      
library(kableExtra) 
library(tinytex)
```


```{r}
## load data
# @Jessica this data path will work if you set your directory to our program eval repo
df_data <- read_csv("ps2_data_spring26.csv", show_col_types = FALSE)
```

```{r}
glimpse(df_data)
summary(df_data)
```


1. 
The ideal experiment would be to randomly assign grid electricity connection 
to SMEs, and find the difference between the average firm annual revenues for 
SMEs who are assigned grid electricity and SMEs who are not assigned. We can 
calculate this difference by using the formula for the average treatment effect, 
which is as follows: 
tau ATE = E[Yi|Di=1] - E[Yi|Di=0]
Di is whether SMEs are assigned to grid electricity connections, with 1 meaning 
the firm is connected and 0 meaning the firm is not connected. Yi is the outcome after being connected to the grid, which in this case would be a firm's annual revenue. E means that we take the average of SMEs' annual revenue across firms. 
We take two averages: the average firm annual revenue for firms who are 
connected to the electricity grid and the average firm annual revenue for firms 
who are not connected to the grid. The tau ATE is the difference in average revenues between SMEs who are connected to the grid and SMEs who are not 
connected.



2.
If we simply compared connected to non-connected firms, then we wouldn't be 
accounting for fundamental differences between the firms, which would distort our 
calculation of the estimated effect of electricity connection. In other words,
we would get the naive estimator, which does not account for selection bias.

The balance table below shows that the connected and non-connected firm groups are 
statistically different (p>0.05) on the number of livestock and the number of 
employees. This systemic difference between the two groups reduces my faith in the
comparison that PEDAP wants to make because it shows that the two groups are 
meaningfully different on factors that could impact revenue.

From ta session 2:
```{r}
# apply the regression to all pre-treatment covariates
regressions <- df_data %>%
  select("owner_older40",
         "owner_male",
         "num_livestock",
         "years_edu",
         "num_employees",
         "grid_subsidy_amount") %>% 
  lapply(function(x) lm(x ~ grid_connection, data = df_data))

# pull the regression coefficient and p-value
balance_table <- regressions %>%
  sapply(function(x) coef(summary(x))[c(2,8)]) %>%
  t()

# format the table output
balance_table %>%
  kable(col.names = c("Differences in means", "p-value")) %>%
  kable_styling(position = "center", font_size = 8.5, latex_options = "hold_position")
```


3. (optional, ungraded)

4. We decided to utilize all available variables because of their potential impact
on revenue, but the tradeoff here is that we couldn't use as many data points due 
to the curse of dimensionality. We saw the curse of dimensionality at play when, 
for example, we had an unconnected farm with 10 employees but no *connected* 
farm with 10 employees available for comparison.
Our estimated average treatment effect (ATE) is 12,231, suggesting that the presence of
grid connection contributes to a $12,231 annual revenue increase, on average, in 
Uganda. This estimated ATE is *less* than the naive estimator (12,992) by about
$700/yr.

Lab 3:
```{r}
# check for common support for the conditional independence assumption
# make control variable
df_data <- df_data %>%
  mutate(no_grid_connection = ifelse(grid_connection==0,1,0))

# are there treated and control people at each level of the X variable?
df_data %>%
  group_by(num_employees) %>%
  summarize(n_connected = sum(grid_connection),
            n_unconnected = sum(no_grid_connection),
            .groups = "keep") # to see total number of employee groups
```

```{r}
# check common support for multi-variable "cells" (check if we have treated and control observations across multiple observable variables)
df_data %>%
  group_by(owner_older40, owner_male, num_livestock, num_employees) %>%
    summarize(n_connected = sum(grid_connection),
            n_unconnected = sum(no_grid_connection),
            .groups = "keep") 

```
```{r}
# find the usable cells that we can find a treatment effect for
usable_cells <- df_data %>%
   group_by(owner_older40, owner_male, num_livestock, num_employees) %>%
    summarize(n_connected = sum(grid_connection),
            n_unconnected = sum(no_grid_connection),
            .groups = "keep") %>%
  filter(n_connected > 0) %>%
  filter(n_unconnected > 0)

usable_cells
```

```{r}
# calculate the treatment effect

# calculate the means by cell * treatment status
grp <- df_data %>%
  group_by(owner_older40, owner_male, num_livestock, num_employees) %>%
  summarize(mean_revenue_connected =
              sum(firm_revenue*grid_connection)/sum(grid_connection),
            mean_revenue_unconnected = sum(firm_revenue*no_grid_connection)/
              sum(no_grid_connection),
            count_connected = sum(grid_connection),
            count_unconnected = sum(no_grid_connection),
            .groups = "keep")

# convert the zero count cells to NA
grp <- grp %>% mutate(mean_revenue_connected = ifelse(count_connected==0, NA,
                                                      mean_revenue_connected),
                      mean_revenue_unconnected = ifelse(count_unconnected==0,NA, 
                                                        mean_revenue_unconnected))

# take difference for each cell
grp$diff <- grp$mean_revenue_connected - grp$mean_revenue_unconnected
grp_fin  <- grp %>% filter(!is.na(diff))
```

```{r}
# estimate tau ATE as the weighted average
ate <- weighted.mean(grp_fin$diff,
                     w = (grp_fin$count_connected + 
                            grp_fin$count_unconnected))
ate
```
```{r}
# calculate treatment effect using the naive estimator
naive <- mean(df_data$firm_revenue[df_data$grid_connection == 1]) - 
  mean(df_data$firm_revenue[df_data$grid_connection == 0])

naive
# estimate tau ATE as a weighted average
ate <- weighted.mean(grp_fin$diff, 
                     w = (grp_fin$count_connected + grp_fin$count_unconnected)) 

ate

# Estimate the tau_att as as weighted average
att <- weighted.mean(grp_fin$diff,
                     w = grp_fin$count_connected) 
att
```

```{r}
# Find naive estimator

mean(subset(df_data, subset = grid_connection==1)$firm_revenue) - mean(subset(df_data, subset = grid_connection==0)$firm_revenue)

```


I decided to include all variables except for years of education and grid 
subsidy amount as these two variables had significant differences in means 
between the treatment and control groups. These characteristics were unbalanced 
between the treatment and control groups which could affect the calculated ATE. 
The calculation finds that SMEs who are connected to the grid have an increased 
12111.93 dollars in firm annual revenue compared to SMEs who are not connected 
to the grid. The strength of this approach is now it is able to account for 
variables covariates to isolate the estimated treatment effect. The weakness of 
this approach is that the more covariates that are included in the calculation 
to be matched with each other, the less likely matches occur and the smaller 
the data set becomes. The result from the exact matching approach (12111.93) 
is lesser than the result from the naive estimator (12992.22). 
5.

6. We would use grid subsidy access as an instrumental variable for grid 
connection status We would first examine the impact of the subsidy on revenue.
Then we would estimate the impact of the subsidy on a farm's likelihood of grid
connection. We'd finally arrive at our estimated IV treatment effect by dividing
the former estimate by the latter.

In math: 
Reduced form (regressing Yi on Zi): Yi = alpha + theta*Zi + πXi + ni
where i is an individual farm, Zi equals 1 if i received the subsidy and
0 otherwise, Xi represents covariates, and ni represents the error term.

First stage (regressing Di on Zi): Di = alpha + gamma*Zi + ßXi + ni
where Di is 1 if i is connected to the grid and 0 otherwise, and Zi, Xi, and ni
have the same meaning as in the previous formula.

IV estimate = theta/gamma


7.

8. This regression estimates the impact of grid subsidy access (not actual grid 
connection) on revenue, a.k.a. the reduced form estimator. This parameter has 
limited use for policy. On the one hand, policymakers should care about how the 
subsidy impacts revenue. On the other hand, selection bias could distort the 
estimate because the subsidy might only be used by farms who would have obtained 
grid connection anyway. Essentially, this is a useful parameter in concept, but 
the aforementioned regression is an unreliable way to estimate it. The regression 
produces a coefficient of -6.6977 (p<0.05), suggesting that the subsidy decreases 
annual revenue by about $7 per year on average.


```{r}
# run regression (reduced form)

reg_subsidy <- lm(df_data$firm_revenue ~ df_data$grid_subsidy_amount + df_data$owner_older40 + df_data$owner_male + df_data$num_livestock + df_data$years_edu + df_data$num_employees + df_data$grid_connection)

summary(reg_subsidy)

```

9. 

10. Random noise associated with the subsidy amount data would constitute a classical
measurement error in Di. This causes attenuation bias (it biases the causal 
estimate towards 0) because of the bias that is embedded in the treatment (see
Lecture 8, slides 16-17). Luckily, the use of instrumental variables can 
correct the bias. 

A systemic error in the subsidy amount data (non-classical measurement error in
Di) would also bias our estimate, and instrumental variables could not address
the problem.






