---
title: "ProgramEvalPS1"
author: "Jessica Prus and Arianna Wooten"
date: "2026-04-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stargazer)  
library(broom)      
library(kableExtra) 
library(tinytex)
```

```{r}
## load data
df_data <- read_csv("C:/Users/blueb/Downloads/ps1_data_spring26.csv", show_col_types = FALSE)
```

```{r}
glimpse(df_data)
summary(df_data)
```


1.


2. Measuring household specific treatment effects is impossible because of 
selection bias. The households who receive and don't receive the microcredit 
might be systematically different from each other, would would in turn affect 
the magnitude of the results which may not represent the true impact. 

3.


4. In this context, the ATT would be the average outcome from receiving the 
microcredit on households who did receive the microcredit subtracted by the 
average outcome from not receiving the microcredit on the households who did 
receive the microcredit. The FARMS data will not allow us to estimate the ATT, 
because we are unable to observe the outcomes of households who received the 
microcredit had they not received the microcredit. 


5.



6.
```{r}
# create histogram of post treatment incomes

# define an object that holds earnings for treated households
treated_inc <- df_data$endline_income[df_data$farms_trt == 1]

# compute the median treated income
med_treat_inc <- median(treated_inc, na.rm = TRUE)

# plot income of treated households
ggplot(data.frame(treated_inc), aes(x = treated_inc)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 10) +
  geom_vline(xintercept = med_treat_inc, color = "red") +
  annotate("text",
           x = med_treat_inc,
           y = Inf, 
           label = paste("Median =", round(med_treat_inc, 2)), 
           vjust = 2, 
           color = "red") +
  labs(title = "Post-Treatment Income on Treated Households",
       x = "Income",
       y = "Number of Households") +
  theme_minimal()

# define an object that holds earnings for untreated households
untreated_inc <- df_data$endline_income[df_data$farms_trt == 0]

# compute the median untreated income
med_untreat_inc <- median(untreated_inc, na.rm = TRUE)

# plot income of untreated households
ggplot(data.frame(treated_inc), aes(x = treated_inc)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 10) +
  geom_vline(xintercept = med_untreat_inc, color = "red") + 
  annotate("text",
           x = med_untreat_inc,
           y = Inf, 
           label = paste("Median =", round(med_untreat_inc, 2)), 
           vjust = 2, 
           color = "red") +
  labs(title = "Post-Treatment Income on Unreated Households",
       x = "Income",
       y = "Number of Households") +
  theme_minimal()

```

The difference in average income between the treatment and control group is 
4721 - 1378 = 3343.

```{r results='asis', warning=FALSE}
# estimate the ATE

base_rg <- lm(endline_income ~ farms_trt,
              data = df_data)

summary(base_rg)
```

The interpretation of the coefficient is that access to the microcredit 
increased income by 2736.63 units. This result is statistically 
significant. This is 606.37 units less of an increase than the simple 
difference in means. The effect is larger relative to the mean in the control 
group (1378).

7.


8.


9.


10.



