---
title: "Problem Set 2"
author: "Jessica Prus and Arianna Wooten"
date: "2026-05-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(tidyverse)
library(stargazer)  
library(broom)      
library(kableExtra) 
library(tinytex)
library(AER)
```


```{r}
## load data
df_data <- read_csv("ps2_data_spring26.csv", show_col_types = FALSE)
```

```{r}
glimpse(df_data)
summary(df_data)
```


1. The ideal experiment would be to randomly assign grid electricity connection 
to SMEs, and find the difference between the average firm annual revenues for 
SMEs who are assigned grid electricity and SMEs who are not assigned. We can 
calculate this difference by using the formula for the average treatment effect, 
which is as follows: 
tau ATE = E[Yi|Di=1] - E[Yi|Di=0]
Di is whether SMEs are assigned to grid electricity connections, with 1 meaning 
the firm is connected and 0 meaning the firm is not connected. Yi is the outcome 
after being connected to the grid, which in this case would be a firm's annual 
revenue. E means that we take the average of SMEs' annual revenue across firms. 
We take two averages: the average firm annual revenue for firms who are 
connected to the electricity grid and the average firm annual revenue for firms 
who are not connected to the grid. The tau ATE is the difference in average 
revenues between SMEs who are connected to the grid and SMEs who are not 
connected.



2. If we simply compared connected to non-connected firms, then we wouldn't be 
accounting for fundamental differences between the firms, which would distort 
our calculation of the estimated effect of electricity connection. In other 
words, we would get the naive estimator, which does not account for selection 
bias.

The balance table below shows that the connected and non-connected firm groups 
are statistically different (p>0.05) on the number of livestock and the number 
of employees. This systemic difference between the two groups reduces my faith 
in the comparison that PEDAP wants to make because it shows that the two groups 
are meaningfully different on factors that could impact revenue.

```{r}
# apply the regression to all pre-treatment covariates
regressions <- df_data %>%
  select("owner_older40",
         "owner_male",
         "num_livestock",
         "years_edu",
         "num_employees",
         "grid_subsidy_amount") %>% 
  lapply(function(x) lm(x ~ grid_connection, data = df_data))

# pull the regression coefficient and p-value
balance_table <- regressions %>%
  sapply(function(x) coef(summary(x))[c(2,8)]) %>%
  t()

# format the table output
balance_table %>%
  kable(col.names = c("Differences in means", "p-value")) %>%
  kable_styling(position = "center", font_size = 8.5, 
                latex_options = "hold_position")
```


3. (optional, ungraded)

4. We decided to exclude the number of employees and number of livestock because 
their distribution was not significantly different across the treated and 
untreated groups. We saw the curse of dimensionality at play when certain 
grid-connected firms had no matching non-connected firms available for 
comparison. Our estimated average treatment effect (ATE) is 12,565, suggesting 
that the presence of grid connection contributes to a $12,565 annual revenue 
increase, on average, in Uganda. This estimated ATE is *less* than the naive 
estimator (12,992) by about $450/yr.

```{r}
# check for common support for the conditional independence assumption
# make control variable
df_data <- df_data %>%
  mutate(no_grid_connection = ifelse(grid_connection==0,1,0))

# are there treated and control people at each level of the X variable?
df_data %>%
  group_by(num_employees) %>%
  summarize(n_connected = sum(grid_connection),
            n_unconnected = sum(no_grid_connection),
            .groups = "keep") # to see total number of employee groups
```

```{r}
# check common support for multi-variable "cells" (check if we have treated and control observations across multiple observable variables)
df_data %>%
  group_by(owner_older40, owner_male, num_livestock, num_employees) %>%
    summarize(n_connected = sum(grid_connection),
            n_unconnected = sum(no_grid_connection),
            .groups = "keep") 

```
```{r}
# find the usable cells that we can find a treatment effect for
usable_cells <- df_data %>%
   group_by(owner_older40, owner_male, years_edu, grid_subsidy_amount) %>%
    summarize(n_connected = sum(grid_connection),
            n_unconnected = sum(no_grid_connection),
            .groups = "keep") %>%
  filter(n_connected > 0) %>%
  filter(n_unconnected > 0)

usable_cells
```

```{r}
# calculate the treatment effect

# calculate the means by cell * treatment status
grp <- df_data %>%
  group_by(owner_older40, owner_male, years_edu, grid_subsidy_amount) %>%
  summarize(mean_revenue_connected =
              sum(firm_revenue*grid_connection)/sum(grid_connection),
            mean_revenue_unconnected = sum(firm_revenue*no_grid_connection)/
              sum(no_grid_connection),
            count_connected = sum(grid_connection),
            count_unconnected = sum(no_grid_connection),
            .groups = "keep")

# convert the zero count cells to NA
grp <- grp %>% mutate(mean_revenue_connected = ifelse(count_connected==0, NA,
                                                      mean_revenue_connected),
                      mean_revenue_unconnected = ifelse(count_unconnected==0,NA, 
                                                        mean_revenue_unconnected))

# take difference for each cell
grp$diff <- grp$mean_revenue_connected - grp$mean_revenue_unconnected
grp_fin  <- grp %>% filter(!is.na(diff))
```

```{r}
# estimate tau ATE as the weighted average
ate <- weighted.mean(grp_fin$diff,
                     w = (grp_fin$count_connected + 
                            grp_fin$count_unconnected))
ate
```

```{r}
# calculate treatment effect using the naive estimator
naive <- mean(df_data$firm_revenue[df_data$grid_connection == 1]) - 
  mean(df_data$firm_revenue[df_data$grid_connection == 0])

naive
# estimate tau ATE as a weighted average
ate <- weighted.mean(grp_fin$diff, 
                     w = (grp_fin$count_connected + grp_fin$count_unconnected)) 

ate

# Estimate the tau_att as as weighted average
att <- weighted.mean(grp_fin$diff,
                     w = grp_fin$count_connected) 
att
```

```{r}
# Find naive estimator

mean(subset(df_data, subset = grid_connection==1)$firm_revenue) - 
  mean(subset(df_data, subset = grid_connection==0)$firm_revenue)

```
I decided to include all variables except for years of education and grid 
subsidy amount as these two variables had significant differences in means 
between the treatment and control groups. These characteristics were unbalanced 
between the treatment and control groups which could affect the calculated ATE. 
The calculation finds that SMEs who are connected to the grid have an increased 
12111.93 dollars in firm annual revenue compared to SMEs who are not connected 
to the grid. The strength of this approach is now it is able to account for 
variables covariates to isolate the estimated treatment effect. The weakness of 
this approach is that the more covariates that are included in the calculation 
to be matched with each other, the less likely matches occur and the smaller 
the data set becomes. The result from the exact matching approach (12111.93) 
is lesser than the result from the naive estimator (12992.22).


5. Using instrumental variables (IV) would be a plausible strategy if the first 
stage (land gradients are related to being assigned a grid connection) and 
exclusion restriction (if land gradients only affect firm revenue through grid connections) requirements are met. The limitation of using an IV instrument is 
that it may be difficult to prove the exclusion restriction as it involves the 
error term of the equation. The exclusion restriction is satisfied by developing 
a convincing argument that the IV instrument only affects the outcome variable 
through the treatment. An alternative instrumental variable could be the type 
of cable that links SMEs to the national power grid. Power cables with higher 
voltage can affect the length of time required for charging or utilizing 
machinery, in turn affecting the firm's productivity and outputs which affects 
firm revenue. Power cables only affect firm revenue through connection to the 
grid. This would be a good instrument because power cable voltage remains 
constant and if provided by the government, does not face issues of spillovers 
or inaccurate reporting by SMEs.


6. We would use grid subsidy access as an instrumental variable for grid 
connection status We would first examine the impact of the subsidy on revenue.
Then we would estimate the impact of the subsidy on a firm's likelihood of grid
connection. We'd finally arrive at our estimated IV treatment effect by dividing
the former estimate by the latter.

In math: 
Reduced form (regressing Yi on Zi): Yi = alpha + theta*Zi + πXi + ni
where i is an individual firm, Zi equals 1 if i received the subsidy and
0 otherwise, Xi represents covariates, and ni represents the error term.

First stage (regressing Di on Zi): Di = alpha + gamma*Zi + ßXi + ni
where Di is 1 if i is connected to the grid and 0 otherwise, and Zi, Xi, and ni
have the same meaning as in the previous formula.

IV estimate = theta/gamma


7.
```{r}
# run first stage regression of grid subsidy amount on grid connection
first_stage <- lm(grid_connection ~ grid_subsidy_amount + years_edu, 
                  data=df_data)

summary(first_stage)
```
This estimates the variation in grid connection take-up caused by the grid 
subsidy amount. The estimate finds that 0.0015452 percentage points of the 
variation in grid connection take-up is caused by the grid subsidy amount. 


8. This regression estimates the impact of grid subsidy access (not actual grid 
connection) on revenue, a.k.a. the reduced form estimator. This parameter has 
limited use for policy. On the one hand, policymakers should care about how the 
subsidy impacts revenue. On the other hand, selection bias could distort the 
estimate because the subsidy might only be used by firms who would have obtained 
grid connection anyway. Essentially, this is a useful parameter in concept, but 
the aforementioned regression is an unreliable way to estimate it. The 
regression produces a coefficient of 12.90 (p<0.05), suggesting that the 
subsidy increases annual revenue by about $12.90 per year on average.

```{r}
# run regression (reduced form)

reg_subsidy <- lm(df_data$firm_revenue ~ df_data$grid_subsidy_amount + df_data$owner_older40 + df_data$owner_male + df_data$num_livestock + 
                    df_data$years_edu + df_data$num_employees)

summary(reg_subsidy)

```


9. 
```{r}
# run first stage regression of grid subsidy amount on grid connection
first_stage <- lm(grid_connection ~ grid_subsidy_amount + years_edu, 
                  data=df_data)

# run second stage regression of grid connection on firm revenue
second_stage <- lm(firm_revenue ~ first_stage$fitted.values + years_edu, 
                   data = df_data)

summary(second_stage)
```

```{r}
# run iv reg using ivreg
iv_reg <- AER::ivreg(firm_revenue ~ grid_connection + years_edu | 
                       grid_subsidy_amount + years_edu, data = df_data)
summary(iv_reg)
```

```{r}
# alternate method (non-2SLS):
# I also messed around with the covariates here FYI

# Step 1: calculate the reduced form (as in Question 8)

reg_subsidy <- lm(df_data$firm_revenue ~ df_data$grid_subsidy_amount + df_data$owner_older40 + df_data$owner_male + df_data$num_livestock + 
                    df_data$years_edu + df_data$num_employees 
                   )

summary(reg_subsidy) # reduced form est = 12.90

# Step 2: calculate the first stage estimate (as in Question 9)
first_stage <- lm(grid_connection ~ grid_subsidy_amount + years_edu + owner_older40 + owner_male + num_livestock + years_edu + num_employees, 
                  data=df_data)

summary(first_stage) # first stage estimate = 0.0016333

# Step 3: Divide reduced form / first stage

12.90 / 0.0016333 # final estimate: 7898.12 
  
```
Grid electricity connection does matter for firm revenues. Being connected to 
the electricity grid increases firm revenues by $7409.7. This finding is 
statistically significant, with a p-value of <2e-16. The standard error is 
596.4, which does not cause concern because it is small relative to the 
coefficient estimate of 7409.7 and it's smaller than the residual standard 
deviation of 3258.


10. Random noise associated with the subsidy amount data would constitute a 
classical measurement error in Di. This causes attenuation bias (it biases the 
causal estimate towards 0) because of the bias that is embedded in the treatment 
(see Lecture 8, slides 16-17). Luckily, the use of instrumental variables can 
correct the bias. 

A systemic error in the subsidy amount data (non-classical measurement error in
Di) would also bias our estimate, and instrumental variables could not address
the problem.






