---
title: "Problem Set 2"
author: "Jessica Prus and Arianna Wooten"
date: "2026-04-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(tidyverse)
library(stargazer)  
library(broom)      
library(kableExtra) 
library(tinytex)
```


```{r}
## load data
# @Jessica this data path will work if you set your directory to our program eval repo
df_data <- read_csv("ps2_data_spring26.csv", show_col_types = FALSE)
```

```{r}
glimpse(df_data)
summary(df_data)
```


1. 
The ideal experiment would be to randomly assign grid electricity connection 
to SMEs, and find the difference between the average firm annual revenues for 
SMEs who are assigned grid electricity and SMEs who are not assigned. We can 
calculate this difference by using the formula for the average treatment effect, 
which is as follows: 
tau ATE = E[Yi|Di=1] - E[Yi|Di=0]
Di is whether SMEs are assigned to grid electricity connections, with 1 meaning 
the firm is connected and 0 meaning the firm is not connected. Yi is the outcome after being connected to the grid, which in this case would be a firm's annual revenue. E means that we take the average of SMEs' annual revenue across firms. 
We take two averages: the average firm annual revenue for firms who are 
connected to the electricity grid and the average firm annual revenue for firms 
who are not connected to the grid. The tau ATE is the difference in average revenues between SMEs who are connected to the grid and SMEs who are not 
connected.



2.
If we simply compared connected to non-connected firms, then we wouldn't be 
accounting for fundamental differences between the firms which would distort our 
calculation of the estimated effect of electricity connection.

from ta session 2
```{r}
# apply the regression to all pre-treatment covariates
regressions <- df_data %>%
  select("owner_older40",
         "owner_male",
         "num_livestock",
         "years_edu",
         "num_employees",
         "grid_subsidy_amount") %>% 
  lapply(function(x) lm(x ~ grid_connection, data = df_data))

# pull the regression coefficient and p-value
balance_table <- regressions %>%
  sapply(function(x) summary(coef(x))[c(2,8)]) %>%
  t()

# format the table output
balance_table %>%
  kable(col.names = c("Differences in means", "p-value")) %>%
  kable_styling(position = "center", font_size = 8.5, latex_options = "hold_position")
```


3. (optional, ungraded)

4.
```{r}
# check for common support for the conditional independence assumption
# make control variable
df_data <- df_data %>%
  mutate(no_grid_connection = ifelse(grid_connection==0,1,0))

# are there treated and control people at each level of the X variable?
df_data %>%
  group_by(num_employees) %>%
  summarize(n_connected = sum(grid_connection),
            n_unconnected = sum(no_grid_connection),
            .groups = "keep") # to see total number of employee groups
```
```{r}
# check common support for multi-variable "cells" (check if we have treated and control observations across multiple observable variables)
df_data %>%
  group_by(owner_older40, owner_male, num_livestock, num_employees) %>%
    summarize(n_connected = sum(grid_connection),
            n_unconnected = sum(no_grid_connection),
            .groups = "keep") 

```
```{r}
# find the usable cells that we can find a treatment effect for
usable_cells <- df_data %>%
   group_by(owner_older40, owner_male, num_livestock, num_employees) %>%
    summarize(n_connected = sum(grid_connection),
            n_unconnected = sum(no_grid_connection),
            .groups = "keep") %>%
  filter(n_connected > 0) %>%
  filter(n_unconnected > 0)

usable_cells
```

```{r}
# calculate the treatment effect

# calculate the means by cell * treatment status
grp <- df_data %>%
  group_by(owner_older40, owner_male, num_livestock, num_employees) %>%
  summarize(mean_revenue_connected =
              sum(firm_revenue*grid_connection)/sum(grid_connection),
            mean_revenue_unconnected = sum(firm_revenue*no_grid_connection)/
              sum(no_grid_connection),
            count_connected = sum(grid_connection),
            count_unconnected = sum(no_grid_connection),
            .groups = "keep")

# convert the zero count cells to NA
grp <- grp %>% mutate(mean_revenue_connected = ifelse(count_connected==0, NA,
                                                      mean_revenue_connected),
                      mean_revenue_unconnected = ifelse(count_unconnected==0,NA, 
                                                        mean_revenue_unconnected))

# take difference for each cell
grp$diff <- grp$mean_revenue_connected - grp$mean_revenue_unconnected
grp_fin  <- grp %>% filter(!is.na(diff))
```

```{r}
# estimate tau ATE as the weighted average
ate <- weighted.mean(grp_fin$diff,
                     w = (grp_fin$count_connected + 
                            grp_fin$count_unconnected))
ate
```
```{r}
# calculate treatment effect using the naive estimator
naive <- mean(df_data$firm_revenue[df_data$grid_connection == 1]) - 
  mean(df_data$firm_revenue[df_data$grid_connection == 0])

naive
```


I decided to include all variables except for years of education and grid 
subsidy amount as these two variables had significant differences in means 
between the treatment and control groups. These characteristics were unbalanced 
between the treatment and control groups which could affect the calculated ATE. 
The calculation finds that SMEs who are connected to the grid have an increased 
12111.93 dollars in firm annual revenue compared to SMEs who are not connected 
to the grid. The strength of this approach is now it is able to account for 
variables covariates to isolate the estimated treatment effect. The weakness of 
this approach is that the more covariates that are included in the calculation 
to be matched with each other, the less likely matches occur and the smaller 
the data set becomes. The result from the exact matching approach (12111.93) 
is lesser than the result from the naive estimator (12992.22). 
5.

6.

7.

8.

9.

10.




