---
title: "ProgramEvalPS1"
author: "Jessica Prus and Arianna Wooten"
date: "2026-04-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stargazer)  
library(broom)      
library(kableExtra) 
library(tinytex)
```

```{r}
## load data
# @Jessica this data path will work if you set your directory to our program eval repo
df_data <- read_csv("ps1_data_spring26.csv")

#df_data <- read_csv("C:/Users/blueb/Downloads/ps1_data_spring26.csv", show_col_types = FALSE)

```

```{r}
glimpse(df_data)
summary(df_data)
```
1. The impact of access to microcredit (Yi) for a household i on income is the difference between the household's income with and without access to microcredit. 

In math: Ti = Yi(1) - Yi(0),
where Ti is the income impact for household i, Yi(1) is the income of household i with access to microcredit, and Yi(0) is the income of household i without access to microcredit.

2. Measuring household specific treatment effects is impossible because of 
selection bias. Selection bias is relevant because the households who receive 
and don't receive the microcredit might be systematically different from each 
other, would would in turn affect the magnitude of the results which may not 
represent the true impact. Measuring household specific treatment effects is 
also impossible due to the fundamental problem of causal inference as we are 
unable to observe both potential outcomes of a household being treated and 
untreated at the same time. We are unable to calculate the counterfactual, and 
are therefore unable to calculate individual household effects.


3. This estimator measures not just the average treatment effect on income, but the income impact of other variables that distinguish households with microcredit access from households without microcredit access. Their suggested estimator is sometimes called the "naive estimator" because of the selection problem; access to microcredit is likely not provided at random, which means that there are other factors that may bias the estimate, unequally impacting income for those with access to microcredit and those without. For example, households with access to microfinancing might tend to have more resourced local governments, which assist household income with a variety of interventions in addition to microcredit options. This disparity, among others, would bias the estimate.


4. ATT = E[Yi(1) - Yi(0)|Di=1] 
In this context, the ATT would be the average outcome from receiving the 
microcredit on households who did receive the microcredit subtracted by the 
average outcome from not receiving the microcredit on the households who did 
receive the microcredit. The FARMS data will not allow us to estimate the ATT, 
because we are unable to observe the outcomes of households who received the 
microcredit had they not received the microcredit. 

5. The histograms suggest that treated households have a higher average income pre-treatment. The balance table shows that the treated and untreated groups were statistically different on income, agricultural expenditure, household size, and livestock ownership even before treatment (access to microcredit for the treatment group). In order to estimate the causal effect of farms_trt, we must be able to reasonably assume that the treatment and control group are statistically the same on unobserved characteristics.

```{r}
# HISTOGRAMS
# create histogram of pre-treatment incomes

# define an object that holds baseline income for treated households
treated_inc_pre <- df_data$baseline_income[df_data$farms_trt == 1]

# plot income of treated households
ggplot(data.frame(treated_inc_pre), aes(x = treated_inc_pre)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 10) +
  labs(title = "Pre-Treatment Income on Treated Households",
       x = "Income",
       y = "Number of Households") +
  theme_minimal()

# repeat for untreated households
untreated_inc_pre <- df_data$baseline_income[df_data$farms_trt == 0]

# plot income of untreated households
ggplot(data.frame(untreated_inc_pre), aes(x = untreated_inc_pre)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 10) +
  labs(title = "Pre-Treatment Income on Untreated Households",
       x = "Income",
       y = "Number of Households") +
  theme_minimal()
```
```{r}
# BALANCE TEST
# run regression on each covariate (week 2 TA session)

regressions <- df_data %>% 
  select("baseline_income", 
         "baseline_agexp", 
         "baseline_hhsize", 
         "baseline_livestock") %>% 
  lapply(function(x) lm(x ~ farms_trt, data = df_data))

# pull the regression coefficient and p-value
balance_table <- regressions %>%
  sapply(function(x) coef(summary(x))[c(2,8)]) %>%
  t()

# format the table output
balance_table %>%
  kable(col.names = c("Differences in means", "p-value")) %>%
  kable_styling(position = "center", font_size = 8.5, latex_options = "hold_position")

```

6.
```{r}
# create histogram of post treatment incomes

# define an object that holds earnings for treated households
treated_inc <- df_data$endline_income[df_data$farms_trt == 1]

# compute the median treated income
med_treat_inc <- median(treated_inc, na.rm = TRUE)

# plot income of treated households
ggplot(data.frame(treated_inc), aes(x = treated_inc)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 10) +
  geom_vline(xintercept = med_treat_inc, color = "red") +
  annotate("text",
           x = med_treat_inc,
           y = Inf, 
           label = paste("Median =", round(med_treat_inc, 2)), 
           vjust = 2, 
           color = "red") +
  labs(title = "Post-Treatment Income on Treated Households",
       x = "Income",
       y = "Number of Households") +
  theme_minimal()

# define an object that holds earnings for untreated households
untreated_inc <- df_data$endline_income[df_data$farms_trt == 0]

# compute the median untreated income
med_untreat_inc <- median(untreated_inc, na.rm = TRUE)

# plot income of untreated households
ggplot(data.frame(treated_inc), aes(x = treated_inc)) +
  geom_histogram(fill = "skyblue", color = "white", bins = 10) +
  geom_vline(xintercept = med_untreat_inc, color = "red") + 
  annotate("text",
           x = med_untreat_inc,
           y = Inf, 
           label = paste("Median =", round(med_untreat_inc, 2)), 
           vjust = 2, 
           color = "red") +
  labs(title = "Post-Treatment Income on Unreated Households",
       x = "Income",
       y = "Number of Households") +
  theme_minimal()

```

The difference in average income between the treatment and control group is 
4721 - 1378 = 3343.

```{r results='asis', warning=FALSE}
# estimate the ATE

base_rg <- lm(endline_income ~ farms_trt,
              data = df_data)

summary(base_rg)
```

The interpretation of the coefficient is that access to the microcredit 
increased income by 2736.63 units. This result is statistically 
significant. This is 606.37 units less of an increase than the simple 
difference in means. The effect is larger relative to the mean in the control 
group (1378).

7.


8. In light of this new information, I estimated the intent-to-treat (ITT) 
estimate in question 6. The ITT estimate measures the effect of offering 
treatment, which includes non-compliance. The ITT differs from the ATE in that 
the ATE measures the actual effect of the treatment while the ITT measures the 
effect of offering treatment. The ITT is always less than the ATE.


9.


10. If they are correct, then the estimate in question 6 underestimates the 
treatment effect. If the control households have increased income as a result 
of their neighbors receiving the microcredit, then the true difference between 
the effects of receiving the microcredit and not receiving the microcredit is 
higher than the calculated difference from question 6. The income of control 
households is an overestimate, which results in downward bias for the estimated 
treatment effect. The estimated treatment effect in question 6 is the lower 
bound for the true treatment effect because spillovers make the average outcome 
higher than without treatment.




